%!TEX root = ../thesis.tex
% Chapter --- Background

\chapter{Background} \label{sec:background}

This section sets out the theory the project rests on. It starts with the clinical motivation (mammographic screening and the two classification tasks), then builds up the feature-extraction side: why handcrafted texture features are useful, what wavelets are, and how the wavelet scattering transform turns wavelets into a stable feature representation. It then covers, more briefly, convolutional neural networks and the idea of fusing the two feature types, since both are part of the wider pipeline this project replicates. The aim is to explain the ideas, not to re-derive them; precise implementation choices and parameter values are given later in Section~\ref{sec:methodology}.

\section{Mammography and breast cancer screening}

Breast cancer is the most commonly diagnosed cancer in women worldwide \cite{sung2021global}, and X-ray mammography is the primary imaging tool used to screen for it. A mammogram is a low-dose X-ray projection of the breast; the goal of screening is to find suspicious changes early, while they are still small and more treatable. Large-scale reviews of screening programmes estimate that inviting women for regular mammographic screening reduces breast cancer mortality by roughly 20\%, albeit at the cost of some overdiagnosis \cite{marmot2012benefits}. In practice a radiologist reads each mammogram and records findings using a standardised vocabulary --- the American College of Radiology's BI-RADS lexicon, which encodes both an overall suspicion category and a breast-density category \cite{sickles2013acr}.

Reading mammograms is time-consuming and subject to inter-reader variability, which has motivated a long line of computer-aided diagnosis (CAD) research that tries to support the radiologist with automated image analysis \cite{burt2018deep}. This project, following Razali et al.\ \cite{razali2023cnn}, looks at two distinct classification problems that arise in this setting. The first is \textbf{tissue-density classification}: deciding whether a region of breast tissue is fatty or fibroglandular (dense). Density matters clinically because dense tissue both raises cancer risk and makes tumours harder to see \cite{boyd2007mammographic}, and it is one of the BI-RADS categories \cite{sickles2013acr}. The second is \textbf{mass classification}: deciding whether a detected mass lesion is benign or malignant. These two tasks are different in character --- density is a property of relatively uniform background texture, whereas a mass is a localised structured object --- and this distinction recurs throughout the report.

\section{Feature extraction and texture in mammograms}

Before any classifier can be trained, an image (or image patch) has to be turned into a numerical feature vector. Broadly there are two families of approach. \emph{Handcrafted} features are computed with a fixed, human-designed procedure; \emph{learned} features are optimised from data by a neural network (Section~\ref{sec:cnns-bg}). Handcrafted features have a particular appeal in medical imaging because they need no training data, are deterministic, and are easier to reason about --- useful when annotated datasets are small, as they typically are for mammography.

For mammograms, much of the discriminative information lives in \emph{texture}: the spatial pattern of intensity variation rather than the overall brightness. Fatty and fibroglandular tissue differ mainly in how their intensity fluctuates across the image, and benign and malignant masses differ in the regularity and sharpness of their internal structure and margins. A classical way to quantify texture is the grey-level co-occurrence matrix (GLCM) \cite{haralick1973textural}, which summarises how often pairs of intensities occur at given offsets. A limitation of single-scale texture descriptors is that breast structures appear at many sizes at once --- from fine trabecular patterns to larger glandular regions --- so a representation that captures structure at \emph{multiple scales} is better matched to the problem. This multi-scale requirement is what motivates wavelet-based features.

\section{Wavelets and multi-scale analysis}

A wavelet is a small, localised oscillating waveform. By dilating (stretching) and rotating a single mother wavelet, one obtains a family of filters tuned to different scales and orientations \cite{morlet1982wave}. This multi-scale (multiresolution) decomposition --- representing a signal as a coarse approximation plus added detail at successively finer scales --- is the foundation of wavelet image analysis \cite{mallat1989theory}. Convolving an image with such a filter bank measures how much ``edge-like'' or ``oscillatory'' content the image has at each scale and orientation, and at each location. This is the basic idea of multi-scale image analysis: coarse wavelets respond to large, smooth structures, while fine wavelets respond to small, sharp ones.

Wavelets are well suited to texture and medical image analysis for two reasons. First, because they are localised in space (unlike the global sinusoids of the Fourier transform), they are stable under small geometric deformations --- a slightly warped texture produces only a slightly changed wavelet response \cite{mallat2012invariant}. Second, the multi-scale, multi-orientation decomposition naturally matches the way biological texture is organised across scales. The drawback is that a raw wavelet transform is \emph{not} translation invariant: shifting the image shifts all the coefficients, so two patches of the same tissue type at different positions look different to a classifier. Removing this nuisance variability, while keeping the useful texture information, is exactly what the scattering transform is designed to do.

\section{The wavelet scattering transform} \label{sec:wst-bg}

The wavelet scattering transform (WST), introduced by Mallat \cite{mallat2012invariant} and Bruna and Mallat \cite{bruna2013invariant}, builds a feature representation that is locally invariant to translation and stable to small deformations, while still preserving high-frequency texture information. It does this by cascading three simple operations:

\begin{enumerate}
    \item \textbf{Wavelet convolution.} The signal $x$ is convolved with an oriented wavelet $\psi_\lambda$ (indexed by a scale and orientation $\lambda$), giving $x \star \psi_\lambda$. This is the same multi-scale filtering described above.
    \item \textbf{Modulus non-linearity.} The complex wavelet response is passed through the modulus $|x \star \psi_\lambda|$. Taking the modulus removes the oscillatory phase and produces a smooth, non-negative ``envelope'' of the wavelet response. A non-linearity is essential: without it, the subsequent averaging step would simply destroy the information, because wavelets integrate to zero \cite{bruna2013invariant}.
    \item \textbf{Local averaging.} The envelope is averaged against a low-pass filter $\phi_J$ at scale $2^J$, i.e.\ $|x \star \psi_\lambda| \star \phi_J$. This averaging is what creates local translation invariance: shifts smaller than the window $2^J$ leave the averaged value almost unchanged.
\end{enumerate}

Averaging discards the finer detail of the envelope, but that lost detail is not thrown away --- it is recovered by applying the wavelet--modulus step \emph{again} to the envelope before averaging, giving a second ``layer''. Iterating in this way defines the \textbf{scattering orders}. The zeroth order is just the smoothed image $S_0 x = x \star \phi_J$; the first order is $S_1 x = |x \star \psi_{\lambda_1}| \star \phi_J$; the second order is $S_2 x = ||x \star \psi_{\lambda_1}| \star \psi_{\lambda_2}| \star \phi_J$; and so on. Concatenating the coefficients of all orders gives the scattering representation. (The precise definitions and parameters used in this project are given in Section~\ref{sec:ws}.)

Two theoretical properties make this representation attractive. \textbf{Translation stability:} the averaging by $\phi_J$ makes the descriptor invariant to translations small relative to the scale $2^J$, so the same tissue at different positions maps to nearly the same features \cite{mallat2012invariant}. \textbf{Deformation stability:} because wavelets are localised and the modulus is a smooth non-linearity, a small deformation of the image produces only a small, bounded change in the scattering coefficients (a Lipschitz-continuity property) \cite{mallat2012invariant, bruna2013invariant}. In practice this means the WST is robust to exactly the kinds of small geometric variation seen between mammogram patches of the same class. Bruna and Mallat further show that most of the representation's energy is captured within the first two orders, with negligible energy beyond order two, which is why second-order scattering is the standard choice \cite{bruna2013invariant}. Crucially, the wavelet filters are fixed in advance --- the WST has no parameters to learn --- so it can be applied directly to small datasets.

\section{Scattering features for classification}

The scattering transform outputs coefficient \emph{maps}: for each path (the chain of scales and orientations defining $S_0$, $S_1$, $S_2$, \ldots) there is a small spatially-subsampled image of averaged values. To obtain a fixed-length feature vector suitable for a standard classifier, these maps are reduced to scalars --- most simply by \emph{spatially averaging} each map, giving one number per path. Stacking those numbers produces the feature vector that is fed to a classifier. This is the representation used in the original scattering papers \cite{bruna2013invariant} and in the reference paper for this project \cite{razali2023cnn}.

Once an image is represented as such a vector, any general-purpose classifier can be used, because the hard work of building an informative, invariant representation has already been done by the transform. Standard choices include PCA-based generative classifiers, kernel support vector machines, logistic regression, and nearest-neighbour methods \cite{bruna2013invariant, cover1967nearest}; the reference paper uses a Subspace $k$-nearest-neighbours ensemble, which combines many $k$-NN classifiers each trained on a random subset of features \cite{ho1998random, cover1967nearest, pedregosa2011scikit}. The key point is that the classifier sits \emph{on top of} the fixed scattering features and can be chosen independently of them; the small-sample rationale for this ensemble choice is given in Section~\ref{sec:classifier}.

How the coefficient maps are summarised is itself a design choice with several options. \emph{Spatial averaging} (one scalar per path) is the most invariant and compact, and is appropriate when the patch is treated as a homogeneous texture. \emph{Preserving the coefficient maps} (keeping the spatial grid, or richer per-path statistics) retains information about \emph{where} structure occurs, which can matter for localised objects but reduces translation invariance. A third option is the \emph{scattering covariance} of Cheng and M\'enard \cite{cheng2021scattering}, which augments the averaged coefficients with correlation statistics computed \emph{between} scattering paths. This captures higher-order dependencies in the texture --- information that single-path averages miss --- while still producing a compact set of scalar descriptors rather than a flattened spatial tensor. This project uses spatial averaging for its baseline scattering features and explores the scattering covariance as a richer alternative (Sections~\ref{sec:ws} and~\ref{sec:scov_tissue}).

\section{Convolutional neural networks} \label{sec:cnns-bg}

A convolutional neural network (CNN) is the learned counterpart to the handcrafted pipeline above \cite{lecun1998gradient}. Like the WST it applies banks of convolutional filters across the image, but its filters are \emph{learned} from labelled data by gradient descent rather than fixed in advance, so stacked layers progress from simple local patterns to increasingly complex, task-specific structure before a final classifier head. Deep CNNs such as the residual network (ResNet) architecture \cite{he2016deep} achieve strong results across medical imaging tasks \cite{burt2018deep}.

The trade-off relative to the scattering transform is essentially data versus flexibility. A CNN can, in principle, learn whatever features best separate the classes, but it needs a large amount of labelled data to do so and can overfit when data is scarce --- a real concern for mammography datasets, which are small. The WST, by contrast, has no learnable filters: it cannot adapt to the data, but it also cannot overfit, and it works directly on small datasets. The two approaches are therefore complementary, which is the motivation for combining them.

\section{Fusion of scattering and CNN features}

\emph{Fusion} means combining the two feature types into a single classification decision, in the hope of getting the best of both: the data-driven flexibility of the CNN and the stable, training-free texture description of the scattering transform. There are two broad strategies. In \textbf{feature-level fusion}, the feature vectors from each branch are concatenated into one longer vector that a single classifier is trained on. In \textbf{decision- (or model-) level fusion}, each branch produces its own prediction and these are combined afterwards, for example by averaging probabilities or majority voting. The reference paper performs feature-level fusion, concatenating CNN and scattering features before the final classifier, and reports that the fused representation outperforms either branch alone \cite{razali2023cnn}. The intuition is that the scattering features supply robust texture information that helps most precisely when the CNN is data-starved and at risk of overfitting. Ensemble and combination methods more generally are a well-established way to improve classifier robustness \cite{ho1998random}.

\section{Why spatial information matters for mass lesions}

Wavelet scattering buys its stability through \emph{spatial averaging}: each coefficient map $S_n(\vec{x})$ is collapsed to a single scalar $\langle S_n \rangle$ per path, which makes the descriptor invariant to small translations and deformations. For a roughly homogeneous texture --- such as a background-tissue patch in the fatty-versus-fibroglandular task --- this is exactly the right inductive bias: the discriminative signal is the \emph{statistics} of the texture, not where any particular structure sits, so averaging discards only nuisance position information.

A mass lesion is different. Malignancy is encoded less in a single homogeneous texture and more in the \emph{spatial arrangement} of distinct sub-regions: a dense lesion \emph{core}, the \emph{margin} or boundary (spiculated or irregular margins are a classical malignancy cue, whereas smooth circumscribed margins suggest benignity), and the immediately \emph{surrounding} parenchyma (architectural distortion, tethering). Spatially averaging the scattering maps over the whole patch blends these regions together and erases precisely the boundary-versus-core contrast that separates benign from malignant masses. This is consistent with the empirical pattern reported later (Chapter~\ref{sec:results}): averaged wavelet scattering is a strong tissue descriptor but a weak mass descriptor.

This observation motivates the project's main mass-side extension. Rather than discard \emph{where} the lesion structure lies, we (i)~keep the scattering \emph{maps} at a lower invariance scale ($J=3$, which retains a $28\times28$ spatial grid per coefficient channel) and let a small convolutional head learn over them, and (ii)~explicitly decompose each lesion into core/inner, boundary, and surrounding-context regions before scattering, so the descriptor is conditioned on lesion morphology (Chapter~\ref{sec:extensions}). Tissue-density classification, being texture-like, does not need this spatial conditioning and is instead improved on the WS side by a richer \emph{scalar} descriptor, the scattering covariance.

\section{Relation to this project}

The pipeline replicated and extended in this project follows directly from the ideas above. Mammograms are first preprocessed (normalisation, pectoral-muscle removal, breast masking, contrast enhancement) so that later feature extraction sees clean, comparable inputs. Fixed-size patches are then extracted for the two tasks: background tissue patches for fatty-versus-fibroglandular density classification, and mass-lesion patches for benign-versus-malignant classification. Each patch is passed through the wavelet scattering transform (and, as an extension, the scattering covariance) to produce handcrafted texture features, and separately through a fine-tuned CNN to produce learned features. The two feature types are evaluated on their own and then fused, with a classifier trained on top to make the final tissue or mass prediction. The remainder of the report makes each of these steps concrete (Section~\ref{sec:methodology}) and reports how well they reproduce the reference paper's results (Section~\ref{sec:results}).


% ============================================================
% 3. METHODOLOGY
% ============================================================
