%!TEX root = ../thesis.tex
% Chapter --- Methodology

\chapter{Data, Preprocessing, Methods, and Experimental Protocol} \label{sec:methodology}

This section describes the end-to-end pipeline: from loading and preprocessing the raw mammographic data, through patch extraction and tissue labelling, to wavelet scattering feature computation and classification. The pipeline is implemented across several Jupyter notebooks, each corresponding to one stage, and the full codebase is available in the accompanying project repository.

\section{Dataset: INbreast} \label{sec:dataset}

We use the INbreast dataset \cite{moreira2012inbreast}, a publicly available full-field digital mammography (FFDM) database comprising 410 DICOM images acquired at a breast centre in Porto, Portugal. Each case includes standard cranio-caudal (CC) and medio-lateral oblique (MLO) views.

Three complementary metadata sources are linked by a common numeric \texttt{file\_id}:

\begin{itemize}
    \item \textbf{INbreast.xls} --- per-image metadata including BI-RADS assessment category \cite{sickles2013acr}, ACR breast density score (1--4), laterality, and acquisition view.
    \item \textbf{OsiriX XML annotations} --- plist-format files (343 of 410 images) containing region-of-interest (ROI) polygons for lesions such as masses.
    \item \textbf{Pectoral muscle masks} --- LabelMe-format JSON polygon annotations \cite{aliniya2023pectoral} covering 202 MLO-view images, providing ground-truth segmentations for the pectoral muscle.
\end{itemize}

For the tissue classification task, a binary density label is derived from the ACR score: ACR 1--2 maps to non-dense (class 0), and ACR 3--4 maps to dense (class 1). For mass classification, the BI-RADS assessment determines the label: BI-RADS 2--3 is labelled benign, and BI-RADS 4--6 is labelled malignant; BI-RADS 1 (normal) is excluded.

This binary mapping yields \textbf{409 of the 410 DICOMs} as usable for the tissue task. One row drops out at ACR cleanup: \texttt{file\_id 53582540}, a BIRADS-1 normal right-MLO image whose \texttt{ACR} field in \texttt{INbreast.xls} is a whitespace-only string (a data-entry artefact in the original spreadsheet) and therefore cannot be placed on the binary density scale. The dropped image carries no mass annotation, no lesion ROI, and no OsiriX XML file, so the exclusion has no downstream consequence for the mass classification task or any other stage. This step and the identity of the dropped row are recorded in \textit{Step 0.2 --- Map to XLS}, \S 7.1.

The resulting tissue-task class distribution is moderately imbalanced: \textbf{282 non-dense} (ACR~1: 136; ACR~2: 146) vs.\ \textbf{127 dense} (ACR~3: 99; ACR~4: 28), i.e.\ approximately a \textbf{69\,\%/31\,\%} split. The majority-class baseline against which any tissue classifier should be compared is therefore around \textbf{69\,\%}: a model that always predicts non-dense would achieve this. Read in that light, the WS-only baseline of ${\sim}84\%$ test accuracy in Section~\ref{sec:stage1} sits roughly \textbf{15~pp above} the majority baseline, and the adapted scattering covariance result of ${\sim}95.9\%$ test (Section~\ref{sec:scov_tissue}) sits roughly \textbf{27~pp above} it. Downstream classifiers do not apply class-weight balancing; metrics are reported on the natural distribution.


\section{Preprocessing} \label{sec:preprocessing}

Each DICOM image passes through a four-stage preprocessing pipeline, implemented in \textit{Step 1.1 --- Preprocessing}:

\begin{enumerate}
    \item \textbf{Normalisation.} The pixel array is read using pydicom \cite{pydicom2024} and rescaled to the interval $[0, 1]$. The reference paper converts each DICOM to PNG (an implicit min-max windowing into 8-bit) and then normalises to $[0, 1]$; we approximate this by clipping each image at its 1st and 99th percentiles and then rescaling. A direct min-max stretch was tested as an ablation and was found to compress the breast-tissue dynamic range into a narrow band near zero, because a small number of pixels per image (calibration markers, edge artefacts, sensor hot spots) sit far above the bulk distribution of breast tissue and dominate the rescaling. The percentile clip is robust to those extreme values, keeps the meaningful tissue values well-spread across $[0, 1]$, and improved 10-fold CV accuracy on the fatty-vs-fibroglandular task from ${\sim}80\%$ to ${\sim}83\%$ on this dataset. This is a small implementation adaptation, not a methodological change. A header audit of all 410 INbreast DICOMs (Step 0.1) further confirmed that every file uses \texttt{MONOCHROME2} photometric interpretation and that none carry VOI LUT or \texttt{WindowCenter}/\texttt{WindowWidth} metadata. Earlier versions of the loader applied a VOI LUT and inverted \texttt{MONOCHROME1} images defensively; both branches were no-ops on this dataset and have now been removed.

    \item \textbf{Pectoral muscle removal (MLO/ML views only).} For oblique-view images with a matching ground-truth JSON mask, the pectoral muscle region is zeroed out. This prevents the pectoral muscle from being mistaken for dense breast tissue during subsequent analysis. Two independent pectoral-mask sources exist for INbreast (the LabelMe JSON files used here, and an Apple-plist XML set from the original INESC Porto release); the choice has only a small downstream effect, which is audited in Section~\ref{sec:annot_audit}. Coverage of the LabelMe set is \textbf{202 of the $\sim$206 MLO/ML views}: 4 MLO/ML DICOMs have no pectoral mask in this source and silently bypass the pectoral step (any small pectoral region they contain is then handled by the Otsu breast-mask step below), and 1 CC view (\texttt{file\_id 50994273}) carries an anomalous small polygon that the loader's MLO/ML view-gating filters out.

    \item \textbf{Breast mask segmentation.} The breast region is isolated using Otsu thresholding \cite{otsu1979threshold} on the uint8-converted image, followed by morphological closing and opening with an 11-pixel elliptical kernel, extraction of the largest connected component, and hole filling. This produces a binary mask separating the breast from the background.

    \item \textbf{Contrast enhancement.} Contrast-Limited Adaptive Histogram Equalization (CLAHE) \cite{pizer1987adaptive} is applied within the breast mask using a clip limit of 2.0 and a tile grid size of $8 \times 8$, via OpenCV \cite{bradski2000opencv}. This locally enhances tissue contrast while preventing noise amplification.
\end{enumerate}

\textbf{Storage format.} Preprocessed images and the extracted mass / background patches are saved as NumPy \texttt{.npy} arrays rather than the PNG files the reference paper uses. We checked that this choice does not materially affect downstream results: a controlled comparison on the 115-patch WS mass-classification task --- feeding the Subspace $k$-NN ensemble with WS features extracted from \texttt{.npy} patches vs.\ an 8-bit PNG round-trip of the same patches (float $[0,1] \rightarrow \mathrm{uint8} \rightarrow$ 8-bit grayscale PNG $\rightarrow$ reload $\rightarrow$ float $[0,1]$) --- produced \emph{identical} held-out test accuracy on all five train/test split seeds tested ($\{1, 7, 34, 42, 99\}$), with cross-validation accuracy and AU-ROC differences well inside the across-seed noise floor (e.g.\ ${\sim}0.2$~pp on CV-mean vs.\ a ${\sim}2$~pp across-seed standard deviation, with the two non-zero deltas pointing in opposite directions on different metrics). At the pixel level the 8-bit PNG round-trip introduced a per-pixel MAE of ${\sim}0.08\%$ --- a half-step of 8-bit quantisation, as expected --- which the Wavelet Scattering transform's spatial-averaging step then absorbs, leaving a feature-vector relative MAE of only ${\sim}0.02\%$. We therefore retain \texttt{.npy} as a practical implementation choice (no quantisation step, exact preservation of mask dtypes, and the native format expected by Kymatio, PyTorch and scikit-learn), not as an accuracy improvement. The full numerical evidence is in \texttt{data/outputs/png\_vs\_npy\_investigation/}.

\textbf{Differences from the reference paper.} Table~\ref{tab:preproc_comparison} summarises the key differences. The most consequential is contrast enhancement: the paper uses a density-aware histogram adjustment (their reference [73], by the same authors) that adapts to the clinical breast-density category, whereas our standard CLAHE is density-agnostic --- likely one source of the performance difference on the density task, since it forgoes exactly the density-specific contrast being classified. The remaining differences (PNG conversion; seeded region growing versus ground-truth polygon masks) are tabulated rather than expanded here.

\begin{table}[h]
\centering
\caption[Comparison of the preprocessing pipeline: paper versus replication]{Preprocessing pipeline comparison between the reference paper and this replication.}
\label{tab:preproc_comparison}
\begin{tabular}{p{3.2cm}p{5.5cm}p{5.5cm}}
\toprule
\textbf{Step} & \textbf{Paper \cite{razali2023cnn}} & \textbf{This replication} \\
\midrule
Normalisation & DICOM $\rightarrow$ PNG (implicit min-max windowing); rescale to $[0,1]$ & 1st/99th-percentile clip; rescale to $[0,1]$ \\
\addlinespace
Image format & DICOM $\rightarrow$ PNG conversion & Operate directly on DICOM arrays \\
\addlinespace
Background removal & Morphological operations; background set to NaN (MATLAB convention) & Otsu threshold + morphology + largest component + hole filling; background set to 0 \\
\addlinespace
Pectoral muscle removal & Seeded region growing (MLO views) & Ground-truth LabelMe JSON polygon masks \cite{aliniya2023pectoral} (202 MLO views) \\
\addlinespace
Contrast enhancement & Density-aware histogram adjustment [73] (adapts to ACR category) & CLAHE (clip limit = 2.0, tile grid $8 \times 8$); density-agnostic \\
\addlinespace
Patch tissue labelling & Otsu threshold on remaining breast region & Otsu threshold + ACR-guided class filtering \\
\bottomrule
\end{tabular}
\end{table}


\section{Patch Extraction and Tissue Labelling} \label{sec:patch_extraction}

Background tissue patches for the fatty versus fibroglandular classification task are extracted in \textit{Step 1.3 --- Background Patch Sampling}. The procedure operates on the 343 images that have XML annotations:

\begin{enumerate}
    \item \textbf{Mass exclusion.} All ROIs labelled ``Mass'' in the XML annotations are parsed into polygon masks and dilated by 24 pixels to create an exclusion zone, ensuring that sampled background patches do not overlap with lesion tissue.

    \item \textbf{Tissue segmentation.} Within the remaining allowed region (breast mask minus pectoral muscle minus dilated mass), an Otsu threshold splits tissue into a darker fatty region and a brighter fibroglandular region. A small gap of 0.03 around the threshold is enforced to reduce ambiguous patches at the boundary.

    \item \textbf{ACR-guided sampling.} Unlike the paper, which does not describe density-aware patch filtering, we use the ACR breast density score to restrict which tissue classes each image may contribute. Images with ACR 1 (almost entirely fatty) contribute only fatty patches; ACR 4 (extremely dense) contributes only fibroglandular patches; and ACR 2--3 contribute both. This prevents mislabelling that can occur when Otsu thresholding forces a binary split on a breast that is clinically almost entirely one tissue type.

    \item \textbf{Random patch sampling.} Up to 3 patches per allowed class per image are sampled at random positions and sizes (proportional to the breast bounding box, ranging from 14\% to 30\% of the shorter dimension, capped between 160 and 512 pixels). Each candidate patch must satisfy: full containment within the allowed mask, at least 65\% of pixels belonging to the target tissue class, and a minimum pixel standard deviation of 0.035 to reject flat or near-empty patches. Overlapping patches within the same image are forbidden. All random choices (patch position and size) are drawn from a single seeded generator (\texttt{np.random.default\_rng(34)}), so the sampled patch set is fully deterministic and identical on every run.

    \item \textbf{Resizing.} Accepted patches are resized to $224 \times 224$ pixels using area-based interpolation and saved as float32 NumPy arrays.
\end{enumerate}

This procedure yields 434 background patches in total: 251 fatty and 183 fibroglandular. The reference paper reports 397 patches (210 fatty, 187 fibroglandular); the difference arises from our ACR-guided sampling strategy and the absence of a hard-coded stopping criterion. Table~\ref{tab:patch_counts} compares the patch counts side-by-side.

\begin{table}[h]
\centering
\caption[Number of extracted patches: paper versus this replication]{Number of extracted patches. The paper uses $n = 112$ source images from INbreast; our pipeline processes all 343 images with XML annotations.}
\label{tab:patch_counts}
\begin{tabular}{llcc}
\toprule
\textbf{Task} & \textbf{Class} & \textbf{Paper \cite{razali2023cnn}} & \textbf{This replication} \\
\midrule
\multirow{2}{*}{Background tissue} & Fatty            & 210 & 251 \\
                                   & Fibroglandular   & 187 & 183 \\
\midrule
\multirow{2}{*}{Mass}              & Malignant        & 76  & 75 \\
                                   & Benign           & 36  & 41 \\
\midrule
                                   & \textbf{Total}   & \textbf{509} & \textbf{550} \\
\bottomrule
\end{tabular}

\vspace{0.3em}
{\small Mass patches are used in the benign-vs-malignant task reported in Section~\ref{sec:results}. The BI-RADS mapping used is 2--3 $\rightarrow$ benign and 4--6 $\rightarrow$ malignant; the paper does not state its exact cut-off, which accounts for the 41/36 difference in benign counts.}
\end{table}


\section{Wavelet Scattering Transform} \label{sec:ws}

The wavelet scattering transform provides a mathematically principled, non-learned feature representation that is locally invariant to translations and stable to small deformations \cite{mallat2012invariant, bruna2013invariant}. For an input signal $x$, the zeroth-order scattering coefficient is the local average $S_0 x = x \star \phi_J$, where $\phi_J$ is a low-pass filter at scale $2^J$. The first-order coefficients are obtained by convolving with a bank of oriented wavelets $\psi_{\lambda_1}$ (parametrised by scale and orientation), applying the modulus non-linearity, and averaging:
\[
S_1 x = |x \star \psi_{\lambda_1}| \star \phi_J.
\]
The second-order coefficients capture higher-frequency information lost by the first-order modulus, computed as:
\[
S_2 x = ||x \star \psi_{\lambda_1}| \star \psi_{\lambda_2}| \star \phi_J.
\]
Concatenating $S_0$, $S_1$, and $S_2$ produces the scattering feature vector.

\textbf{Implementation.} We use the Kymatio library \cite{andreux2020kymatio} with the following configuration, chosen to align with the reference paper:

\begin{itemize}
    \item Maximum scale: $J = 6$, corresponding to an invariance scale of $2^J = 64$ pixels.
    \item Number of orientations: $L = 5$ (Morlet wavelets \cite{morlet1982wave}).
    \item Maximum scattering order: 2.
    \item Input shape: $224 \times 224$.
\end{itemize}

Kymatio returns scattering coefficient \emph{maps} of shape $(C, H_s, W_s)$; following the paper's approach, we spatially average each coefficient map to obtain one scalar per scattering path, yielding a compact 406-dimensional feature vector per patch.

\textbf{Justification of the order and invariance-scale choices.} Two of these settings are directly supported by the foundational scattering analysis of Bruna and Mallat \cite{bruna2013invariant}. First, the choice of maximum order 2 is not arbitrary: they show that scattering energy decays exponentially with path length, with over 99\% of the energy carried by paths of length $m \le 3$, and that second-order coefficients recover precisely the higher-moment texture information that the first-order (SIFT-like) coefficients discard --- enough, for example, to discriminate textures with identical Fourier power spectra. Orders beyond 2 add negligible energy at substantial computational cost, so order 2 is the standard and well-motivated choice. Second, the large invariance scale ($2^J = 64$ on $224 \times 224$ patches) is appropriate because the fatty-versus-fibroglandular task is essentially a \emph{texture} discrimination problem: Bruna and Mallat show that for stationary-texture classification the optimal invariance scale approaches the image (here, patch) width, since this minimises the variance of the windowed-scattering estimator. Treating each background patch as a texture sample therefore favours a large $J$, and the near-complete spatial pooling that results (Kymatio returns a $3 \times 3$ spatial grid per path at this setting) is consistent with that view rather than a loss of useful information.

\textbf{Feature count discrepancy.} The paper reports 391 features using MATLAB's \texttt{waveletScattering} toolbox, which uses 6 orientations. The count follows the formula $1 + JL + \binom{J}{2} L^2 = 1 + 6 \times 6 + 15 \times 36 = 577$; however, the MATLAB toolbox applies $L=6$ orientations and $J_{\max}=5$ wavelet scales (since one scale is reserved for the low-pass filter), giving $1 + 5 \times 6 + \binom{5}{2} \times 6^2 = 391$. In Kymatio with $L = 5$ orientations and $J = 6$ scales (all 6 used for wavelets), the count is $1 + 6 \times 5 + \binom{6}{2} \times 5^2 = 406$. This minor difference is an unavoidable consequence of different library conventions. Table~\ref{tab:ws_config} compares the two configurations.

\begin{table}[h]
\centering
\caption[Wavelet scattering configuration: paper versus this replication]{Wavelet scattering configuration: reference paper (MATLAB) vs.\ this replication (Python / Kymatio).}
\label{tab:ws_config}
\begin{tabular}{lcc}
\toprule
\textbf{Parameter} & \textbf{Paper \cite{razali2023cnn}} & \textbf{This replication} \\
\midrule
Library / toolbox         & MATLAB \texttt{waveletScattering} & Kymatio \cite{andreux2020kymatio} \\
Wavelet type              & Morlet                            & Morlet \\
Invariance scale          & 64 ($2^J$ pixels)                 & 64 ($2^J$ pixels) \\
$J$ (max decomposition)   & 6                                 & 6 \\
Effective wavelet scales  & 5 (1 reserved for low-pass)       & 6 (all used) \\
$L$ (orientations)        & 6                                 & 5 \\
Max scattering order      & 2                                 & 2 \\
Input size                & $224 \times 224$                  & $224 \times 224$ \\
Spatial pooling           & Spatial averaging                 & Spatial averaging \\
Output features           & 391                               & 406 \\
Feature standardisation   & None (raw coefficients)           & \texttt{StandardScaler} \\
\bottomrule
\end{tabular}
\end{table}


\section{Classifier: Subspace k-NN Ensemble} \label{sec:classifier}

Following the paper's specification, we use a \textit{Subspace k-Nearest Neighbours Ensemble} \cite{ho1998random, cover1967nearest} with the following hyperparameters:

\begin{itemize}
    \item Number of ensemble learners: 80.
    \item Subspace dimensionality: 190 features (randomly selected per learner from the full 406-dimensional feature space).
    \item Number of neighbours: $k = 1$.
    \item Random seed: 34.
\end{itemize}

Each learner independently projects the data onto a random 190-dimensional subspace, fits a 1-nearest-neighbour classifier, and produces a class probability estimate. The final prediction is the soft vote (average probability) across all 80 learners. This ensemble approach improves robustness by averaging over many diverse views of the high-dimensional feature space. Beyond matching the reference paper, this choice is also well-suited to our small sample sizes: Bruna and Mallat observe that low-capacity / generative classifiers tend to outperform more flexible discriminative classifiers (such as kernel SVMs) when training samples are scarce, the discriminative classifiers only overtaking them as the training set grows \cite{bruna2013invariant}. A simple distance-based ensemble is therefore a reasonable estimator for our tasks, and especially for the small ($n \approx 115$) mass set. Table~\ref{tab:knn_settings} lists the classifier hyperparameters.

\begin{table}[h]
\centering
\caption[Subspace k-NN ensemble classifier settings]{Subspace k-NN ensemble classifier settings. The paper tested several values and selected the optimal ones shown; we adopt the same optimal values.}
\label{tab:knn_settings}
\begin{tabular}{lccc}
\toprule
\textbf{Parameter} & \textbf{Values tested \cite{razali2023cnn}} & \textbf{Paper optimal} & \textbf{Ours} \\
\midrule
Ensemble method   & Subspace, AdaBoost, RUSBoost & Subspace         & Subspace \\
Learner type      & Nearest neighbour, Discriminant & Nearest neighbour & Nearest neighbour \\
Number of learners ($k$) & 40, 60, 80, 100         & 80               & 80 \\
Subspace dimension       & 170, 190, 210            & 190              & 190 \\
Number of neighbours     & ---                       & 1                & 1 \\
\bottomrule
\end{tabular}
\end{table}

The feature matrix is standardised using \texttt{StandardScaler} from scikit-learn \cite{pedregosa2011scikit} (zero mean, unit variance) before classification, as $k$-NN is distance-based and sensitive to feature scaling. Note that the paper does not mention applying standardisation to the WS features before classification; this is a difference in our pipeline that may affect results.

\textbf{Train--test split.} The 434 patches are split 80/20 into training and test sets, stratified by class, with a fixed random seed of 34 for reproducibility. The split is performed at the \emph{image} level (on unique \texttt{file\_id} values) rather than at the patch level, so that all patches extracted from the same mammogram remain together in either the training or the test partition. This yields 344 training patches (199 fatty, 145 fibroglandular) and 90 test patches (52 fatty, 38 fibroglandular). Cross-validation likewise uses scikit-learn's \texttt{StratifiedGroupKFold} \cite{pedregosa2011scikit} with \texttt{file\_id} as the group label. The rationale for adopting image-level splitting, and the issue that prompted the change, is described in Section~\ref{sec:results} (Stage~2).


\section{Mass ROI Extraction} \label{sec:mass_rois}

For the benign versus malignant mass classification task, mass ROIs are obtained from the INbreast XML annotations (the OsiriX plist files described in Section~\ref{sec:dataset}). Each annotation supplies a bounding box for the lesion. The paper refers generically to ``the mass detection algorithm'' but does not specify which detector; we therefore use the dataset's expert annotations directly, which is the standard usage of INbreast. The procedure, implemented in \textit{Step 1.2 --- Mass ROI Extraction}, is:

\begin{enumerate}
    \item Parse the XML bounding box for each ``Mass'' ROI, yielding $(y_0, y_1, x_0, x_1)$ coordinates.
    \item Crop the corresponding region from the already-preprocessed full image at \texttt{data/outputs/preprocessed/final/}, so the mass patch inherits the same percentile-clip normalisation, breast mask, pectoral removal, and CLAHE enhancement applied to the background patches.
    \item Resize the crop to $224 \times 224$ pixels using area-based interpolation, clip values to $[0, 1]$, and save as a float32 NumPy array.
\end{enumerate}

The label is assigned from the BI-RADS assessment: BI-RADS 2--3 $\rightarrow$ benign (class~0), and BI-RADS 4a/4b/4c/5/6 $\rightarrow$ malignant (class~1). This is the standard clinical convention. The paper does not state its exact mapping, which is a minor source of ambiguity. The procedure yields 116 mass patches (75 malignant + 41 benign); the paper reports 112 (76 + 36). A fuller reconciliation of this small discrepancy --- which we attribute mainly to our retention of all \texttt{Name="Mass"} ROIs (including the few images with multiple mass ROIs) and to the paper's apparent inclusion of some asymmetry/distortion findings as ``mass tissue'' rather than to the BI-RADS threshold --- is given in Section~\ref{sec:mass_count_discussion}.

\section{CNN Branch: ResNet18 Fine-Tuning and Feature Extraction} \label{sec:cnn}

The paper's CNN branch (Model~1 in their Table~6) uses ResNet18 \cite{he2016deep} pre-trained on ImageNet, with the final fully connected layer replaced to output two classes. The paper deliberately selects the comparatively shallow ResNet18 over deeper residual networks (ResNet50/101): with only a few hundred training patches per task, a shallower network is less prone to overfitting and to the information loss that accumulates in very deep architectures on small datasets~\cite{yu2019abnormality}. We replicate this in PyTorch via \texttt{torchvision.models.resnet18(weights=IMAGENET1K\_V1)}, with the FC layer redefined as a $512 \to 2$ linear projection. All layers are fine-tuned (no freezing), matching MATLAB's default \texttt{trainNetwork} behaviour. To meet ResNet's 3-channel input requirement, each single-channel patch is replicated three times and normalised with the standard ImageNet mean and standard deviation. No data augmentation is applied, in line with the paper.

The hyperparameters follow Table~4 of the paper exactly:

\begin{itemize}
    \item Optimiser: Adam, learning rate $\eta = 0.001$, $\beta_1 = 0.9$, $\beta_2 = 0.999$.
    \item $L_2$ weight decay: $1 \times 10^{-4}$.
    \item Loss: cross-entropy.
    \item Number of epochs: 60.
    \item Mini-batch size: 80.
\end{itemize}

After fine-tuning, 512-dimensional feature vectors are extracted from the average-pooling layer of the trained network via a forward hook --- i.e.\ from the penultimate layer, before the final FC. These features are then standardised with \texttt{StandardScaler} and classified by the same Subspace k-NN ensemble described in Section~\ref{sec:classifier}, matching the paper's pipeline for Model~1.

\textbf{Validation set for overfitting tracking.} In addition to the 80/20 train--test split, we carve a $\sim$20\% validation set from the training images (image-level) and track its loss and accuracy at each epoch. This is an additional diagnostic not described in the paper, included to track the CNN overfitting expected on a sample this small (a concern the paper itself emphasises); the diagnostic results are presented in Section~\ref{sec:results} (Stage~4 and Stage~5b). The validation patches are excluded from gradient updates but re-included for the subsequent k-NN feature extraction so that the classifier sees the same training set as the paper.


\section{Region-aware WS-map CNN head (extension)} \label{sec:region_aware_method}

This extension, developed for the mass task, keeps the spatial scattering maps and conditions them on lesion morphology rather than collapsing them to a scalar vector. It has four ingredients.

\textbf{Region masks from polygon ROIs.} For each mass, the OsiriX XML \texttt{Name="Mass"} polygon is rasterised to a binary lesion mask on the preprocessed image. From this single outline, four lesion-relative regions are derived by morphological operations: a \emph{full} region (the lesion with a small surrounding margin), an \emph{inner} region (the eroded core), a \emph{boundary} region (a dilated-minus-eroded annulus straddling the margin), and a \emph{context} region (a dilated surround with the lesion removed). The erosion/dilation structuring elements are sized as a fraction of the lesion bounding box, so the decomposition scales with lesion size. Each region mask is applied to the patch, giving four region images per lesion.

\textbf{$J=3$ scattering maps.} Each region image is passed through the same Kymatio scattering transform but at $J=3$ rather than the paper's $J=6$. This retains a $28\times28$ spatial grid of $91$ coefficient channels per region, instead of collapsing to the near-scalar $3\times3$ grid that $J=6$ produces. Stacking the four regions gives a $4\times91$-channel, $28\times28$ scattering-map tensor per lesion that preserves \emph{where} coefficient energy sits within each lesion sub-region.

\textbf{WS-map CNN head.} A small convolutional network is trained directly on these scattering-map tensors: two $3\times3$ convolutional blocks (Conv--BatchNorm--ReLU, $32$ channels), global average pooling, dropout ($0.5$), and a linear classifier, optimised with Adam ($\eta=10^{-3}$, weight decay $3\times10^{-4}$) for $60$ epochs at batch size $24$. The CNN learns spatial relationships across the region maps that a hand-pooled scalar summary discards; this is the main methodological extension of the project.

\textbf{Comparators.} Two controlled comparators accompany the WS-map CNN. (i)~A \emph{classical} region-aware baseline pools each region map to per-channel statistics and classifies the concatenated vector with logistic regression or the Subspace k-NN ensemble, isolating the value of the region decomposition from the value of the learned head. (ii)~An \emph{improved CNN} comparator extracts $512$-d features from a \emph{frozen} ImageNet ResNet18 on a square lesion crop and classifies them with logistic regression --- a stable, fine-tuning-free CNN baseline. Finally, \emph{late} (decision-level) fusion averages the predicted malignancy probabilities of the region-aware WS$\rightarrow$CNN and the improved CNN, in contrast to the paper's feature-level concatenation.


\section{Final evaluation protocol and reporting} \label{sec:eval_protocol}

The development narrative in Section~\ref{sec:results} reports each stage on a single fixed image-level split (seed~34) with group-aware 10-fold cross-validation, reflecting how the experiments were originally run. For the \emph{final} comparison of the headline methods this is not enough on its own: the mass held-out set contains only ${\sim}25$ patches from ${\sim}22$ images (${\sim}24$ once the data-quality filter of Section~\ref{sec:mass_quality_filter} is applied), so a single additional misclassification shifts test accuracy by ${\sim}4$~pp, and a single split can swing by $10$~pp or more between seeds. The final evaluation protocol therefore follows four principles.

\begin{enumerate}
    \item \textbf{Repeated, image-level splits.} Each headline method is evaluated over five repeated stratified train/test splits (seeds $\{1, 7, 34, 42, 99\}$), each grouped by \texttt{file\_id} so that no patches or ROIs from the same mammogram fall in both the training and test partitions. Reporting the spread across seeds, rather than one fixed split, exposes the split sensitivity that a single number conceals.
    \item \textbf{Leakage-free feature extraction.} Where a method uses a trained feature extractor (the CNN branch), the network is \emph{retrained inside each split} on that split's training images only, so the test features come from a model that never saw them.
    \item \textbf{Training-only scaling.} The \texttt{StandardScaler} is fitted on the training fold only and then applied to the test fold, so that no test statistics influence training.
    \item \textbf{AUROC-led reporting under class imbalance.} The mass task is imbalanced ($74$ malignant vs.\ $41$ benign), so raw accuracy is misleading: a trivial ``always malignant'' rule already scores the ${\sim}68\%$ majority baseline. We therefore lead with the area under the ROC curve (AU-ROC), which is threshold-independent and unaffected by class prevalence, and report accuracy, precision, recall, and F1 alongside it. All metrics are quoted as mean~$\pm$~standard deviation across the five seeds.
\end{enumerate}

Five seeds is a deliberate trade-off: enough to reveal the across-seed variability while keeping every headline method on a common, reproducible set of splits. Because all seeds resample the same ${\sim}115$ mass patches, the reported standard deviation should be read as an indicator of split sensitivity on a small dataset, not as a tight confidence interval. Cross-validation is not used as a headline metric: the repeated held-out splits already give an out-of-sample estimate, and cross-validation computed on \emph{learned} CNN features is optimistic because the feature extractor has already seen the inner-validation images. The consolidated comparison under this protocol is produced by the canonical \textit{FINAL Step~2.1--2.5} notebooks.


\textbf{Reproducibility.} All randomness is pinned per run: the Python, NumPy and PyTorch seeds are set together, and wherever a CNN is trained, cuDNN is placed in deterministic mode (\texttt{torch.backends.cudnn.deterministic=True}, \texttt{benchmark=False}) so that repeated runs of the same seed agree. The canonical diagnostic seed is~$34$; the headline numbers additionally average the five image-level seeds $\{1,7,34,42,99\}$. For the learned WS-map CNN head, which is trained on only ${\sim}90$ patches, we report a five-seed mean~$\pm$~standard deviation and corroborate it with a more stable group-aware cross-validation rather than relying on any single split (Chapter~\ref{sec:discussion}).


% ============================================================
% 3. RESULTS
% ============================================================
