%!TEX root = ../thesis.tex
% Chapter --- Discussion

\chapter{Discussion} \label{sec:discussion}

\section{Replication outcomes at a glance}

Under the final five-seed protocol (Tables~\ref{tab:results_tissue}--\ref{tab:results_mass}) all three paper configurations are reproduced for both tasks. On \emph{tissue}, averaged WS reaches AU-ROC $0.895$ (test $0.839$), trailing the paper's reported number by roughly $7$--$8$~pp --- the largest single replication gap --- while the CNN ($0.979$) and CNN~+~WS fusion ($0.988$) are near-saturated and slightly exceed the paper, which we attribute to a cleaner preprocessing pipeline rather than a methodological advance. On \emph{mass}, averaged WS is weak ($0.726$ AU-ROC) and the CNN-only model reproduces the paper's reported values to within a few points on the seed-34 metrics (including the overfitting signature in its training curves). CNN~+~WS fusion is the strongest strict replication model ($0.816$ AU-ROC, test $0.792$), reproducing the paper's central claim that handcrafted WS features help a data-starved CNN on small-sample mass classification. The development-narrative figures in Chapter~\ref{sec:results} are single seed-34 splits and are superseded by these five-seed means where they differ --- most notably the high-variance mass test numbers (e.g.\ the seed-34 WS-only mass test of $0.52$, which rises to a five-seed $0.701$).

Beyond strict replication, the scattering covariance extension (Section~\ref{sec:scov_tissue}) provides a stronger WS-branch representation \emph{on tissue}: at $J=6, L=3$ it reaches AU-ROC $0.984$ / test $0.959$ (versus averaged WS's $0.895$/$0.839$ and the paper's $0.94$/$0.914$), the strongest adopted WS-branch tissue result in the project. On \emph{mass}, however, it gives no meaningful AU-ROC gain ($0.734$ versus the $0.726$ averaged-WS baseline), so it is adopted as a tissue-side extension only; the mass improvement comes instead from preserving the scattering spatial map and learning over it (Section~\ref{sec:region_aware}), not from a richer scalar descriptor. This is positioned as an improvement to the WS branch, not a paper-replication number, and the comparison to Razali is not apples-to-apples (feature definitions, library, preprocessing, patch counts, and CV scheme all differ).

\section{Fusion fidelity: what the CNN+WS model does and does not reproduce}

The CNN~+~WS fusion implemented here is a faithful instance of the paper's Model~5: a feature-level concatenation of the $512$-dimensional CNN average-pool vector and the $406$-dimensional wavelet-scattering vector ($918$-d in total), classified by the same Subspace k-NN ensemble as the single-feature models, with no GLCM branch. The only unspecified choice we fix is to standardise the concatenated vector on training-fold statistics, which is necessary because the CNN and WS halves occupy very different numerical scales and an un-scaled concatenation would let the larger-magnitude half dominate the distance-based classifier.

On the \textbf{tissue} task the fusion reproduces the paper's qualitative ranking --- fusion is the (jointly) strongest model --- but not the large fusion \emph{gain} the paper reports. Razali's CNN~+~WS rises to $93.5\%$ test from a CNN-only baseline of $77.4\%$, a $+16$~pp jump they attribute to handcrafted features compensating an overfitting CNN. In our pipeline the CNN-only tissue model does not overfit (it already reaches ${\sim}98\%$ test; Section~\ref{sec:results}), so fusion has little left to add: our CNN~+~WS ($96.4\%$ test, AU-ROC $0.988$ over five seeds) is statistically close to CNN-only ($97.9\%$ test, AU-ROC $0.979$). The fusion mechanism is thus reproduced only where it is actually needed, and on tissue --- with cleaner preprocessing --- it is not.

On the \textbf{mass} task the paper's central claim \emph{is} reproduced. Here the CNN genuinely overfits (single-seed CV ${\approx}0.99$ against an honest held-out accuracy of ${\sim}0.77$), and adding the training-free WS features lifts the honest result: under the leakage-free five-seed protocol, fusion (AU-ROC $0.816$, test $0.792$) outranks both CNN-only (AU-ROC $0.762$) and WS-only (AU-ROC $0.726$). This is precisely the regime the paper invokes --- wavelet-scattering texture supporting a data-starved CNN --- and is the clearest evidence in this project that fusion helps.

One caveat tempers the comparison to the paper's headline mass numbers. The paper's reported CNN~+~WS mass result ($91.7\%$ test, AU-ROC $1.00$) is a single split on a ${\sim}25$-patch test set; our \emph{cross-validation} reproduces it (${\approx}98\%$, matching the paper's $98\%$), but the honest repeated held-out estimate is lower and noisier. We therefore treat the paper's CV-led fusion headline as optimistic for very small held-out sets, while affirming its qualitative conclusion that fusion benefits the mass task.

\section{Sources of the WS-only tissue accuracy gap}

The single largest discrepancy is the $\sim$7~pp gap on WS-only tissue classification. Three sources are plausible.

\begin{itemize}
    \item \textbf{Contrast enhancement.} The paper uses a density-aware histogram adjustment that the authors describe in earlier work, which adapts to the ACR breast-density category. We use standard CLAHE, which is density-agnostic. The paper's method is likely to produce features that are more discriminative for the very task being measured (tissue density), because it amplifies density-specific contrast.
    \item \textbf{Library implementation.} Kymatio produces 406-dimensional features (with $L = 5$ orientations, $J = 6$ wavelet scales) whereas the paper's MATLAB toolbox produces 391 (with $L = 6$, effectively $J = 5$). This is a fundamental library convention difference; we cannot make the two match without re-implementing the scattering transform from scratch.
    \item \textbf{Patch sampling.} Our ACR-guided sampling yields 434 patches versus the paper's 397, with a slightly different fatty/fibroglandular ratio. ACR guidance prevents ACR-1 images from contributing spurious fibroglandular patches via Otsu thresholding, but it does shift the marginal class distribution slightly.
\end{itemize}

Of these, the contrast enhancement difference is the most plausibly dominant factor: WS features are essentially a textural descriptor of frequency content, and a density-aware contrast operation literally adjusts the contrast of the very signal WS measures.

\textbf{Closing the gap by changing the feature representation.} A separate observation from the scattering-covariance study (Section~\ref{sec:scov_tissue}) is that the WS-only gap can also be closed --- and substantially exceeded --- by enriching the feature representation rather than by changing preprocessing. The adapted scattering covariance at $J=6, L=3$ reaches AU-ROC $0.984$ on the same patches and same preprocessing pipeline as the original WS, $+0.089$ above the original WS's $0.895$ (in seed-34 group-CV accuracy terms --- a secondary metric --- this is $81.2\% \rightarrow 96.2\%$, $+15$~pp). The interpretation is that the standard Mallat $S_0, S_1, S_2$ coefficients undersample the cross-scale and cross-orientation information that scattering covariance exposes via its $C_{01}, C_{11}, P_{00}, P_{11}$ statistics. The contrast-enhancement and library-convention issues above remain real, but they no longer dominate the achievable WS-branch performance for this project.

\section{Lesson learned: image-level vs.\ patch-level splitting in medical imaging}

The most important methodological lesson of the project was uncovered early in the experiments. When multiple patches are extracted per source image, splitting at the patch level allows highly correlated samples to appear in both training and test sets, producing inflated and meaningless test accuracy. In our first CNN run, $98.9\%$ of test patches shared a source mammogram with the training set; the resulting $100\%$ test accuracy was a direct consequence. Image-level splitting (with group-aware cross-validation) is the standard correction in the medical-imaging literature and is now the default choice throughout this codebase. The fact that the paper's wording was ambiguous on this point is worth flagging for anyone attempting a similar replication.

\section{Mass test-set noise}

The held-out mass test set contains only 25 patches across 22 images. A single misclassification moves test accuracy by 4~pp, so a difference of even $5$--$10$~pp between our test estimate and the paper's may not be a real difference in model performance. Cross-validation accuracy is a more stable comparison point at this sample size, and on that metric our CNN-only mass result ($98.9\%$) is close to the paper's ($100\%$) while our WS-only ($71.1\%$) is in fact slightly \emph{better} than the paper's ($69\%$).

\section{Mass patch count discrepancy with the paper} \label{sec:mass_count_discussion}

Our mass dataset contains 116 patches across 107 source mammograms (75 malignant, 41 benign); the paper reports 112 patches (76 malignant, 36 benign), a net difference of $+5$ benign and $-1$ malignant. The BI-RADS mapping is not the source of this gap: both pipelines place BI-RADS 2--3 in the benign class and BI-RADS 4--6 in the malignant class (the standard clinical convention). Instead, two distinct properties of our extraction pipeline account for it.

% NOTE (manual check -- flagged in review): the two example centroid distances in this paragraph are unverified and could not be reproduced from the XML.
%   * file 51049107 ``89 px'': re-derivation gives ~99 px (polygon-vertex centroid) / ~98 px (bbox centre). This file also has THREE Mass ROIs (a near-duplicate pair plus one ~1700 px away), so ``contains two benign ROIs'' is loose.
%   * file 22427840 ``311 px'': re-derivation gives ~355 px (polygon centroid) / ~379 px (bbox centre); 311 px is not reproducible.
%   The qualitative points (near-duplicate pair vs two separate foci) are correct; the pixel figures are illustrative only. Verify/replace before final submission.
\textbf{Effect 1 --- multi-ROI retention.} Eight of our 107 source images contain more than one \texttt{Name="Mass"} ROI in the XML annotation (seven contain two, one contains three). Five of those images are BI-RADS-2 benign and contribute six extra ROIs above one-per-image; three are BI-RADS-5 malignant and contribute three extras. Inspection of the bounding-box centroids reveals that these multi-ROI cases are a mix of two phenomena: (i)~near-duplicate annotations of a single mass --- e.g.\ file 51049107 contains two benign ROIs whose centroids are only $89$~pixels apart on bounding boxes $150$--$240$~pixels wide; and (ii)~genuinely multifocal disease --- e.g.\ file 22427840 (BI-RADS~5) contains two malignant ROIs whose centroids are $311$~pixels apart, consistent with two separate cancer foci in the same breast (multifocality is well documented in high-BI-RADS cases). Our pipeline retains all such ROIs; collapsing to one ROI per image would yield 107 patches (35 benign + 72 malignant) at the cost of dropping genuine multifocal masses.

\textbf{Effect 2 --- ``Mass tissue'' inclusion criterion.} Our \textit{Step 1.2 --- Mass ROI Extraction} extracts only ROIs whose XML \texttt{Name} tag is literally \texttt{Mass}. The INbreast dataset additionally contains nine images whose XML carries only \texttt{Asymmetry} or \texttt{Distortion} ROIs and no \texttt{Mass} ROI: six asymmetry-only and three distortion-only. Both the XML \texttt{Name} and the corresponding XLS \texttt{Mass} column agree that these are not masses; nevertheless, eight of these nine images carry BI-RADS scores in the 4a--5 range (i.e.\ suspicious or confirmed malignancy), so radiologically they may behave like masses even though they have not been annotated as such by the OsiriX reader. The paper does not state whether it accepts these as ``mass tissue''.

\textbf{A plausible reconstruction of the paper's 112.} If the paper takes one Mass ROI per source image (107 patches, 35 benign + 72 malignant) and then adds one BI-RADS-2 distortion image plus four BI-RADS-4/5 asymmetry-or-distortion images, the totals come to 36 + 76 = 112 --- an exact match with their Table~3 counts. This is consistent with two simultaneous editorial choices on the paper's part: (a)~deduplicate multi-ROI images down to one ROI each, and (b)~broaden the ``mass tissue'' definition to include some architectural-distortion and asymmetry findings. The exact subset cannot be identified from the paper text.

\textbf{Why we retain our extraction.} Our criterion is exactly reproducible from the public INbreast XML alone: any reader can re-derive our 116 patches by filtering XML ROIs on \texttt{Name == "Mass"}. The paper's selection, by contrast, requires per-image judgements --- which non-Mass mass-like findings to include and which multi-ROI duplicates to collapse --- that are not documented and therefore not reproducible without manual review. Keeping all \texttt{Mass} ROIs preserves the clinically important multifocal cases at the small cost of carrying a few near-duplicate annotations; because our train/test split is image-level (Section~\ref{sec:stage3}), duplicates from the same image never cross the train/test boundary, so the inflated count adds slight redundancy within the training fold but introduces no leakage. Finally, the absolute size of the difference is well below the noise floor of the mass test set ($n_\mathrm{test} = 25$): the CNN+WS fusion still reproduces the paper's recall and F1 to within $\sim$1~pp regardless of which counting convention is used. For these reasons we have not changed the extractor to match the paper's number exactly.

\section{Data-quality filter --- one entirely-zero mass patch excluded} \label{sec:mass_quality_filter}

During the scattering covariance experiment (extension method described in Section~\ref{sec:results}), a data-quality issue surfaced that had been silently present in every previous mass result. One INbreast annotation --- \textbf{file\_id 50994354}, BI-RADS 4a, labelled malignant, view \texttt{MG\_R\_ML\_ANON} --- has a bounding box at $y \in [806, 1098]$, $x \in [2214, 2483]$ on a $3328 \times 2560$ preprocessed image. The bbox falls entirely outside the breast mask: \emph{zero of its $78\,548$ pixels} are inside the segmented breast region (which itself covers over a million pixels, so the breast itself is intact). Step~1.1 zeros all non-breast pixels before saving the preprocessed array, so when Step~1.2 crops the bbox from the preprocessed image the result is a uniformly-zero $224 \times 224$ patch.

The root cause appears to be an annotation-or-orientation mismatch between the OsiriX XML and the as-loaded DICOM for this single file: for a right-laterality MLO view the breast itself occupies the left side of the image, so a bbox at $x \in [2214, 2483]$ (very far right of a 2560-wide image) is on the empty chest-wall side. No other mass annotation in INbreast shows this pattern.

\textbf{Effect.} The all-zero patch was in the held-out test split under \texttt{random\_state=34}, so it directly affected the reported mass test metrics for all four mass-using notebooks (Steps~2.1, 2.2, 2.3, 2.4). It contributes an essentially zero feature vector to whichever classifier it reaches; a $k$-nearest-neighbour classifier on such a vector lands on whatever class the training set's nearest non-zero patch happens to be.

\textbf{Filter applied.} A defensive one-line filter was added to every mass-using notebook to exclude patches whose maximum pixel value is zero:
\begin{center}
\texttt{mass\_df = mass\_df[mass\_df["patch\_npy"].apply(lambda p: np.load(p).max() > 0)]}
\end{center}
The filter is dynamic, so it would also catch any future analogous patch, not just this one. The on-disk patches and the upstream \texttt{masses/mass\_index.csv} are unchanged --- only the downstream manifests used by the classifier notebooks were updated. This reduces our mass dataset from 116 patches (75 malignant + 41 benign) to \textbf{115 patches (74 malignant + 41 benign)}.

\textbf{Numerical impact.} On the post-filter mass results (24-patch test set), every method's test accuracy improves by 1--3~pp, since the all-zero patch was a guaranteed-wrong prediction for most methods. CV mean and CV std are essentially unchanged because the patch was in the test split, not in any training fold. The qualitative ranking of methods is preserved. The numerical mass figures reported earlier in this document --- the WS-only and CNN+WS rows of Table~\ref{tab:results_mass}, the $n_\mathrm{test} = 25$ in Section~\ref{sec:results} Stage~5a/b, and the mass-specific numbers cited in Sections~\ref{sec:results} and~\ref{sec:discussion} --- were computed before the filter was applied. Re-running Steps~2.1, 2.2, and~2.3 with the now-filtered mass dataset would shift each test number by $\sim$1--3~pp upward without changing the qualitative picture; we have not re-run them here, but the filter has been added to those notebooks so any future execution will automatically use the 115-patch dataset.

A wider Step~1.2 improvement --- adding a bbox-vs-breast-mask overlap check at extraction time, or cropping from the polygon outline rather than the rectangular bbox --- is deferred as a longer-term cleanup; the one-line filter is sufficient for the current scope.

\section{Annotation source audit --- inter-source disagreements within INbreast} \label{sec:annot_audit}

INbreast ships with several overlapping annotation sources. Two of these inform the pipeline directly: the lesion ROIs (used to extract mass patches) and the pectoral muscle outlines (used in preprocessing). For each, two independent versions exist in the public release, and they do not perfectly agree. The full audit, including visual side-by-side comparisons, is in the project notebook \textit{Step 0.3 --- Investigate Annotation Sources}; the summary below records the conclusions reached there.

\subsection*{Lesion annotations --- XML ROIs vs.\ XLS \texttt{Mass} column}

The presence of a mass in each case is recorded independently in two places:

\begin{itemize}
    \item The \texttt{INbreast.xls} metadata file has a \texttt{Mass} column. The radiologist ticked this with \texttt{X} where a mass was reported. This is a per-file binary flag with no spatial information; \textbf{108} files are flagged.
    \item The OsiriX XML files in \texttt{AllXML/} contain ROI polygons. A subset of ROIs carry the tag \texttt{Name="Mass"}, with pixel-coordinate vertex lists drawn by the (separate) annotator who created the XML files; \textbf{107} files have at least one \texttt{Mass} ROI.
\end{itemize}

Of these, \textbf{106} files appear in both sets. The 3 disagreements are:

\begin{itemize}
    \item \texttt{22614097} (BI-RADS~2, L~CC) and \texttt{22614150} (BI-RADS~2, L~MLO): the XLS marks a mass with \texttt{X}, but the XML annotator drew only calcification ROIs (3 and 4 calcifications respectively), no mass polygon. These are the two XLS-only files. Both are also flagged with \texttt{Micros='X'} in the XLS; the most plausible reading is that the case carried a clinically-benign mass that the XML annotator either considered too small to outline or merged into the calcification cluster.
    \item \texttt{24065584} (BI-RADS~4c, L~CC, ACR~4): the XLS lesion columns are all empty, but the XML annotator drew an unambiguous large \texttt{Mass} polygon. This is the one XML-only file. The XML annotator's outline is consistent with the BI-RADS~4c (highly suspicious for malignancy) classification; the XLS row appears to have simply been left blank.
\end{itemize}

The arithmetic checks out: $108_\mathrm{XLS} - 2_\mathrm{XLS\text{-}only} + 1_\mathrm{XML\text{-}only} = 107_\mathrm{XML}$. The mass-extraction pipeline (Step~1.2) takes ROIs from the XML because it requires polygon coordinates to crop patches from; the XLS \texttt{Mass} column alone provides no spatial information. The choice is therefore forced by what the pipeline needs, but on the merits it works in our favour: the two XLS-only files we miss are both BI-RADS~2 (low-acuity benign), while the one XML-only file we gain is BI-RADS~4c on a dense ACR-4 breast --- exactly the kind of high-acuity malignant case the mass classifier benefits from seeing. The net 107-vs-108 difference is a real inter-annotator inconsistency within INbreast itself, not an artefact of our extraction.

\subsection*{Pectoral muscle masks --- GitHub LabelMe-JSON vs.\ Kaggle Apple-plist XML}

The preprocessing pipeline (Section~\ref{sec:preprocessing}) zeroes out the pectoral muscle on MLO/ML views using ground-truth polygon masks. Two public sources of these masks exist:

\begin{itemize}
    \item \textbf{GitHub source} --- LabelMe-format JSON files by Aliniya \cite{aliniya2023pectoral}, covering \textbf{202} MLO views. This is the source used by the current pipeline.
    \item \textbf{Kaggle source} --- Apple-plist XML files distributed with the original INbreast 1.0 release \cite{moreira2012inbreast} (INESC Porto), covering \textbf{201} MLO views.
\end{itemize}

The two sources agree on the file\_ids of \textbf{199} cases, with 3 GitHub-only (\texttt{20587492}, \texttt{20587518}, \texttt{50994273}) and 2 Kaggle-only (\texttt{24055024}, \texttt{53582737}). The GitHub-only \texttt{50994273} is an unusual case: it is a CC view (no pectoral muscle should be visible in principle), and the GitHub annotator nonetheless drew a polygon for a small sliver of muscle in the corner that the original Porto annotators did not annotate.

On the 199 files both sources annotate, the two masks are visibly different in style:

\begin{itemize}
    \item The GitHub masks are coarse \textbf{triangular regional outlines} (typically 3--8 polygon vertices), filling the upper-corner area where the pectoral muscle is.
    \item The Kaggle masks are fine \textbf{muscle-boundary outlines} (typically 20--40 vertices), hugging the actual muscle tissue.
\end{itemize}

For our pipeline's purpose --- zeroing the muscle region before breast-mask + CLAHE in Step~1.1 --- either source works. The GitHub source removes the entire upper-corner region (muscle plus some surrounding background); the Kaggle source removes only the muscle tissue itself. The downstream impact on classification is expected to be small because (i) the Otsu breast mask removes non-breast pixels regardless of which pectoral mask is used, and (ii) the patch sampler in Step~1.3 enforces a minimum target-fraction (\texttt{min\_target\_frac}), which suppresses patches that straddle the muscle edge. We retain the GitHub source as the default for two reasons: it is already wired into Step~1.1, and its slightly more conservative coverage removes a small amount of additional non-breast pixels that the Kaggle outlines leave behind.

\section{Overfitting signatures}

The expected CNN overfitting is borne out by the per-epoch training/validation curves logged by the Stage~4 and Stage~5b notebooks. Training accuracy saturates at $100\%$ within the first 10 epochs in both tasks. For the tissue task, validation accuracy then stays close to training accuracy, which is why the overfit gap is small. For the mass task, validation accuracy oscillates in the $60$--$90\%$ range, mirroring the paper's reported large CV--test gap and supporting their argument that fusion with handcrafted features is necessary for small-sample mass classification.

\section{What the extension shows}

The extension experiments converge on a single message: for the mass task, \emph{preserving spatial and lesion structure in the wavelet-scattering representation matters more than changing the scalar features or the classifier}. The screening of Section~\ref{sec:screening} shows that normalisation, per-path statistics, coarse pooling, feature selection, and even the richer scattering-covariance descriptor move mass AU-ROC only modestly and within the across-seed noise --- clustered near the $0.726$ averaged-WS baseline and rising to ${\sim}0.86$ for the best hand-pooled variant. The one method that breaks out of that band is the full-patch WS$\rightarrow$CNN head ($0.909$), which keeps the $J=3$ spatial maps and learns over them rather than averaging them away. Conditioning those maps on lesion morphology (region-aware WS) lifts the \emph{classical} (logistic-regression) model from $0.801$ to $0.892$; with a learned CNN head the region-aware maps reach $0.946$ (group-aware CV $0.919 \pm 0.012$), above the whole-patch map+CNN ($0.909$). The region ablation shows the all-four-region tensor is the strongest complete configuration, although neighbouring strong configurations overlap within the reported bands. The operative lever is still preserving the spatial map and learning over it, with the region split retained as a principled, interpretable representation that now also gives the strongest complete region-aware result.

The contrast with tissue is instructive. Tissue density is a texture problem that a richer \emph{scalar} descriptor --- scattering covariance --- already solves ($0.984$ AU-ROC), with no need to retain spatial layout. Mass is a structured-object problem, where \emph{where} the discriminative signal sits (core versus margin versus surrounding context) is decisive, so the scalar route plateaus and the spatial route succeeds. Region-aware preservation is therefore the project's main methodological contribution on the mass side. The CNN and fusion methods support that conclusion rather than replace it: the region-aware late fusion is numerically strongest ($0.948$), but it is a decision-level combination of two trained models, not a new representation, and its margin over the region-aware WS$\rightarrow$CNN alone is within seed noise --- so it is reported as the best \emph{number}, while the region-aware representation is the substantive \emph{result}.

\section{Reproducibility: determinism and holdout-vs-CV stability}

Three reproducibility issues shaped the final evaluation protocol (Section~\ref{sec:eval_protocol}).

\textbf{Small-$n$ instability and seed sensitivity.} The mass held-out set is only ${\sim}24$ patches, so a single split can swing by $10$~pp or more between seeds (the seed-34 WS-only mass test of $0.52$ versus a five-seed mean of $0.701$ is the clearest example). Every headline mass method is therefore reported as a five-seed mean~$\pm$~standard deviation over image-level splits, and the residual bands (e.g.\ $\pm0.036$ for the region-aware WS$\rightarrow$CNN) are an honest reflection of the small sample rather than a defect of the method.

\textbf{Determinism of learned models.} For the learned WS-map CNN, Python/NumPy/PyTorch seeds and cuDNN-deterministic flags are pinned, but the model is still trained on only ${\sim}90$ patches. The mitigation is to average five seeds and corroborate with a tight group-aware cross-validation ($0.919 \pm 0.012$), which is far less seed-sensitive than any single held-out split.

\textbf{Optimistic CV for learned features.} Cross-validation computed on \emph{learned} CNN or fusion features is optimistic, because the feature extractor has effectively already seen the inner-validation fold. The CNN-only and fusion CV figures of ${\sim}0.99$ on mass (marked ``(opt.)'' in Table~\ref{tab:results_mass}) are not honest out-of-sample estimates, whereas their five-seed held-out AU-ROCs ($0.762$ and $0.816$) are. For this reason the project leads with AU-ROC under the repeated image-level holdout as the headline mass metric, and uses group-aware CV only as a stability cross-check --- and, for the classical WS methods where no feature is learned, as a like-for-like paper-comparison metric.

\section{Limitations}

\begin{itemize}
    \item \textbf{Small mass sample.} The mass dataset is only $115$ patches (${\sim}24$ held out per split). The final protocol averages five image-level seeds and reports standard deviations to expose this, but the absolute sample remains small, so the mass AU-ROC bands stay wide; this is the dominant limitation on the mass side.
    \item \textbf{Single dataset.} Only INbreast is used. Cross-dataset generalisation to e.g.\ DDSM or MIAS has not been tested.
    \item \textbf{No augmentation.} Following the paper, no augmentation is used. Augmentation is the most obvious lever to mitigate the observed CNN overfitting.
    \item \textbf{StandardScaler.} The paper does not mention applying feature standardisation before the Subspace k-NN; we do (since k-NN is distance-based). This may help our WS results slightly, but the overall qualitative picture (e.g.\ WS-tissue trailing the paper) is unchanged.
    \item \textbf{GLCM branch excluded.} The GLCM branch and the CNN+GLCM/CNN+WS+GLCM fusion configurations (Models~3, 4, 6 in the paper's Table~6) are deliberately left out of scope. The paper itself reports that adding GLCM to CNN+WS yields no change on accuracy, precision, recall or F1 and only a $+0.01$ AU-ROC gain at the cost of additional training time, so the practical impact of this exclusion is small.
    \item \textbf{Mass AU-ROC volatility.} Even averaged over five seeds, mass AU-ROC carries a visible band (replication fusion $0.816 \pm 0.057$; region-aware WS$\rightarrow$CNN $0.946 \pm 0.036$), because the held-out set is tiny and the probability ranking is coarse. The paper's reported $1.00$ fusion AU-ROC is a single ${\sim}25$-patch split and is not a stable target; a larger external mass set would be needed for a tight estimate.
    \item \textbf{Scale not fully optimised for mass.} The scattering-covariance $J$/$L$ tuning was carried out on the \emph{tissue} task (Section~\ref{sec:scov_tissue}). On mass we ran only an $L$-control at $J=6$ (which showed no gain) and adopted $J=3$ maps for the whole-patch and region-aware models, consistent with Bruna and Mallat's \cite{bruna2013invariant} observation that an object-like task favours a \emph{smaller} invariance scale that retains spatial structure. A full mass-specific covariance $J\times L$ grid was not run; the region-aware results, however, indicate the larger lever for mass is preserving spatial structure rather than re-tuning the invariance scale of a scalar descriptor.
\end{itemize}

% ============================================================
% 5. CONCLUSION
% ============================================================
