%!TEX root = ../thesis.tex
% Chapter --- Introduction

\chapter{Introduction}

Breast cancer is the most prevalent malignancy among women worldwide and the most frequently diagnosed cancer globally, with approximately 2.3 million new cases each year \cite{sung2021global}. Early detection through routine mammographic screening substantially improves patient outcomes, yet the interpretation of mammograms remains a challenging and time-consuming task for radiologists. Computer-aided diagnosis (CAD) systems that leverage computer vision have therefore become an active area of research, offering the potential to assist clinicians in detecting and characterising suspicious findings.

Modern CAD approaches overwhelmingly rely on convolutional neural networks (CNNs), which have demonstrated strong performance for a range of medical imaging tasks \cite{burt2018deep}. However, two significant limitations persist. First, CNNs are data-hungry: their supervised training paradigm requires large volumes of annotated data, while publicly available mammography datasets remain relatively small. Second, conventional CNN architectures primarily extract spatial-domain features and may neglect complementary information present in the frequency domain, such as the multi-scale textural patterns that characterise different tissue types.

The wavelet scattering transform (WST) provides an attractive solution to both issues. Introduced by Mallat \cite{mallat2012invariant} and further developed by Bruna and Mallat \cite{bruna2013invariant}, scattering transforms cascade the input through banks of predefined wavelet filters and non-linear modulus operations to produce feature representations that are locally invariant to translations and stable to small deformations. Crucially, the WST requires no training data at all --- it is a fixed mathematical operation --- and it captures rich multi-scale, multi-orientation texture information in the frequency domain.

Razali et al.\ \cite{razali2023cnn} proposed a hybrid framework that fuses CNN features with WST features (and additionally GLCM texture descriptors \cite{haralick1973textural}) for classifying breast tissue in mammograms, demonstrating that this combination outperforms either branch alone. Their method was evaluated on the INbreast dataset \cite{moreira2012inbreast} for two tasks: distinguishing fatty from fibroglandular background tissue, and classifying mass lesions as benign or malignant.

This project aims to replicate the methodology of Razali et al.\ in Python and to reproduce the classification results reported therein. The pipeline is implemented in a modular fashion so that each component --- preprocessing, patch extraction, wavelet scattering features, CNN features, and the fusion classifier --- can be evaluated independently. The work proceeded iteratively, beginning with the wavelet scattering branch on the fatty versus fibroglandular tissue task, extending to the CNN branch on the same task, then repeating both branches on the benign versus malignant mass task, and finally implementing the CNN~+~WS feature fusion that represents the paper's headline contribution. The development was not linear: an early CNN run revealed a data-leakage issue that required revisiting the train--test split methodology for both branches, which is documented as part of the development narrative in Section~\ref{sec:results}. This report compares all three model configurations (WS-only, CNN-only, CNN~+~WS) against the paper's reported Table~6 (tissue) and Table~7 (mass) benchmarks. The GLCM branch reported in the paper (Models 3, 4, 6 of their Table~6) was excluded from the scope of this project on supervisor advice, since the paper itself shows GLCM contributes negligible improvement when combined with the WS branch.

Beyond replication, the project extends the wavelet-scattering branch along two threads. For the \emph{tissue/background} texture task it adopts the \emph{scattering covariance} representation (Cheng \& M\'enard~\cite{cheng2021scattering}; supervisor-suggested), which augments the standard scattering coefficients with cross-scale and cross-orientation correlation statistics while remaining a compact scalar descriptor. For the \emph{mass} task --- where paper-style \emph{averaged} scattering is weak because spatial averaging discards \emph{where} the lesion structure lies --- it instead \emph{preserves} the scattering spatial map at a lower invariance scale ($J=3$) and learns over it with a small WS-map CNN head; a \emph{region-aware} decomposition (lesion core, margin, whole-lesion, and surrounding-context regions) is the interpretable, morphology-conditioned instantiation of this idea. The contribution is not merely higher accuracy but a methodological finding: \emph{scalar} covariance statistics are effective for tissue texture, whereas mass lesions require \emph{spatial preservation} of the scattering map (the region decomposition being a principled, interpretable refinement that performs on par rather than the source of the gain) rather than a change of scalar feature or classifier. A decision-level late fusion of the WS-map CNN with a frozen-CNN comparator is the numerically strongest mass configuration, but it is reported as an ensemble result rather than the core contribution.

\section{Project Goals}

The specific objectives of this project, following the brief set out by the project supervisor, are:

\begin{enumerate}
    \item Load and explore the INbreast clinical dataset (DICOM mammographic images).
    \item Reproduce the primary pipeline from the reference publication in a modular fashion, enabling comparison of results with and without the wavelet scattering transform.
    \item Reproduce the classification results --- including performance metrics and model comparisons --- both (a)~with and (b)~without the scattering wavelet transform.
    \item Investigate wavelet-scattering extensions for the two tasks, including scattering covariance for tissue/background texture and region-aware scattering-map preservation for mass lesions.
\end{enumerate}

\section{Report Structure}

Section~\ref{sec:background} reviews the theoretical background to the project --- mammography and breast cancer screening, handcrafted texture features, wavelets, the wavelet scattering transform and its use for classification, convolutional neural networks, and feature fusion. Section~\ref{sec:methodology} details the dataset, preprocessing pipeline, patch extraction (background tissue and mass ROIs), wavelet scattering and CNN feature extraction branches, and classifier design. Section~\ref{sec:results} presents the paper-style replication results as a chronological development narrative, covering each stage in the order it was carried out and comparing the outcomes against the paper's benchmarks. Section~\ref{sec:extensions} reports the extension experiments: a screening of supporting non-region-aware methods, the scattering covariance representation for the tissue task, and the region-aware wavelet-scattering deep dive for the mass task. Section~\ref{sec:discussion} discusses sources of discrepancy, lessons learned, the data-quality issues uncovered, limitations, and potential improvements. Section~\ref{sec:conclusion} concludes the report with a summary and directions for further work.


% ============================================================
% 2. BACKGROUND
% ============================================================
